---
title: "Genomic Features between Two Range Sets"
output: 
  html_document:
    toc: true
    toc_depth: 2
params:
  data: options
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The value of a particular genomic feature is pre-calcuated for each base site in the human genome, by examining the 100-bp enclosing window centered upon the site in question. For an interval of continuous genomic sites (termed "a range" in Bioconductor), the site-wise values are averaged to the interval level feature value. The interval-level feature values are compared between two input sets with statistical tests, so that the user will be informed as to which feature(s) demonstrate possible statistically significant difference between the two input interval sets.

The whole human genome is differentiated into three region types: Intron, Exon, and Intergene (intergenic region). Around 1/10000 sites proportionally distributed across chromosomes and three region types were randomly sampled, which gave rise to the background distributions of each feature in each region type. In the histograms below, the superimposed density curves correspond to background distributions of Exon (black), Intron (dark gray), and Intergenic region (light gray), respectively.     


```{r compare, echo=F, comment=F, warning=FALSE, dependson=F, message=F}

library(knitr)

df=read.table("fileList1.txt", sep="\t", header=T,stringsAsFactors = F)
options=split(df$Value, df$Key)

mnFeatures<-c(-132.13,-798.53,2.63,3.62,4.18,46.17,1.81,-135.18,-803.57,2.61,3.62,4.19,48.77,1.21,-131.33,-797.97,2.64,3.63,4.18,45.29,3.44)
sdFeatures<-c(14.31,19.11,0.21,0.20,0.18,13.21,30.86,14.70,20.63,0.16,0.17,0.14,13.35,11.15,13.79,19.52,0.22,0.22,0.20,12.52,57.16)
mnFeatures <- matrix(mnFeatures,nc=3,dimnames=list(c('delta_G37','delta_H','Entropy2','Entropy3','Entropy4','gc','HS'),c('introns','exons','intergene')))
sdFeatures <- matrix(sdFeatures,nc=3,dimnames=dimnames(mnFeatures))


#cmpr2in(): Summarize users' interval-level features in graphics and statistically compare the various features between two inputs.
### INPUT stem1 & stem2: stems of annotated file names (most likely the ultimate input file names to the software).
### regionLoc: directory where bed files for three types of genomic regions are stored.
### INPUT mn & sd: mean & sd statistics of 4*3 background distributions (feature4 by region3)
### NOTE currently I adopt the mean feature score (instead of median, 25%, etc.)
cmpr2in <- function(intv.ls2, stemNames, regionLoc,mn=mnFeatures,sd=sdFeatures) {
  filePrefix<-paste(stemNames[2:1], collapse = "_VS_")
  
	intv1 <- intv.ls2[[1]]
	intv2 <- intv.ls2[[2]]
	################## Discern genomic regions of input intervals #################
	library(GenomicRanges)
	gr2 <- lapply(intv.ls2,deriveGRfrItv); 
	names(gr2) <- stemNames
	regions <- c('introns','exons','intergene')
	rFiles <- paste(regionLoc,paste0(regions,'.bed'),sep='/')
	GR3 <- lapply(rFiles,deriveGRfrBed); 
	names(GR3) <- regions
	regionSum <- matrix(0,nr=2,nc=length(regions),dimnames=list(stemNames,regions )  )
	for (input in stemNames) {
		gr <- gr2[[input]]
		for (region in regions) {
			GR <- GR3[[region]]
			ovlappedItvs <- subsetByOverlaps(gr,GR)
			regionSum[input,region] <- 100*round(length(ovlappedItvs)/length(gr),3)	
		}
	}
	regionSum <- cbind(regionSum,totalNum=c(nrow(intv1),nrow(intv2)))
	colnames(regionSum)[1:length(regions)] <- paste0(colnames(regionSum)[1:length(regions)],'(%)')
	################## End discerning #############################################
  subNames <- list(
    energy=c('delta_G37','delta_H'),
    entropy=c(paste0('Entropy',2:4)),
    gc=c('gc'),
    map=c('HS')
  )
	feature4 <- names(subNames)
	subN <- sapply(subNames,length); names(subN) <- feature4
	######################### Plot foreground/background distributions #####################
	subfs.all <- unlist(subNames)
	cmprTbl <- matrix(NA,nr=length(subfs.all),nc=14,
		dimnames=list(subfs.all,c(paste0('Med',1:2),
				paste(rep(c('W','W+','W-'),each=2),rep(c('','p'),3),sep='.' ), 
				paste(rep(c('D','D+','D-'),each=2),rep(c('','p'),3),sep='.' )
			)
		)
	)
	colnames(cmprTbl) <- gsub('\\.$','',colnames(cmprTbl),perl=T)
	for (feature in feature4) {
		vars <- subNames[[feature]]#paste0('Entropy',c(2,3,4))
		col.ls2 <- lapply(intv.ls2,function(x,cols) x[,cols,drop=F], paste(vars,'mean',sep='_') )
		pdf(paste0(filePrefix, ".", feature,'.pdf'))
		plot_a_feature(col.ls2,stemNames,vars,mn,sd)
		dev.off()
		# Statistical comparison of basewise scores through Wilcoxon & KS test
		for (subf in subNames[[feature]]) {
			subf.ls2 <- lapply(col.ls2,function(x,subf) x[,paste0(subf,'_mean')],subf)
			cmprTbl[subf,] <- itvCmpr_a_subf(subf.ls2)
		}		
	}
	write.table(regionSum,paste0(filePrefix,'.regionSum.tsv'),col.names=NA,quote=F,sep='\t')
	write.table(cmprTbl,paste0(filePrefix,'.cmpr.tsv'),col.names=NA,quote=F,sep='\t')
	return(list(regionSum=regionSum, cmprTbl=cmprTbl))
}
deriveGRfrItv <- function(itvMore) {
	require(GenomicRanges)
	itv <- itvMore[,1:3]
	colnames(itv) <- c('chr','start','end')
  gr <- makeGRangesFromDataFrame(itv)
  gr
}
deriveGRfrBed <- function(bedFile) {
	require(GenomicRanges)
	bed <- read.delim(bedFile,head=F)
	colnames(bed) <- c('chr','start','end')
	gr <- makeGRangesFromDataFrame(bed)
	gr
}
#itvCmpr_a_subf(): for a list of two series of subfeature values, return 6*2 statistics values plus two median values.
itvCmpr_a_subf <- function(ls2) {
	dat1 <- ls2[[1]]
	dat2 <- ls2[[2]]
	D.res <- ks.test(dat1,dat2)
	Dp.res <- ks.test(dat1,dat2,alternative='greater') # plus
	Dm.res <- ks.test(dat1,dat2,alternative='less')
	W.res <- wilcox.test(dat1,dat2)
	Wp.res <- wilcox.test(dat1,dat2,alternative='greater')
	Wm.res <- wilcox.test(dat1,dat2,alternative='less')
	stats <- c(W.res$statistic,W.res$p.value,Wp.res$statistic,Wp.res$p.value,Wm.res$statistic,Wm.res$p.value,
		D.res$statistic,D.res$p.value,Dp.res$statistic,Dp.res$p.value,Dm.res$statistic,Dm.res$p.value)
	stats <- c(median(dat1,na.rm=T),median(dat2,na.rm=T),stats)
	stats
}
plot_a_feature <- function(ls2,stemName2,subNames,mn=mnFeatures,sd=sdFeatures) { #col.ls2,c(stem1,stem2),c('Entropy2','Entropy3','Entropy4')
  library(ggplot2)
  library(gridExtra)
  nSub <- ncol(ls2[[1]])
  pS <- vector('list',nSub)
  mnSubfs <- mn[subNames,,drop=F]
  sdSubfs <- sd[subNames,,drop=F]
	for (i in 1:nSub) {
	  ############ Prepare xs & ys for background distributions #########
		if (subNames[1]!='HS') {
			cols <- c('gray50','black','gray80') #introns,exons,intergene
			names(cols) <- colnames(mn)
			mnSubf <- mnSubfs[i,,drop=F]
			sdSubf <- sdSubfs[i,,drop=F]
			datNorm <- vector('list',ncol(mn)); names(datNorm)=colnames(mn)
			for (region in colnames(mn)) {
				mn.r <- mnSubf[1,region]
				sd.r <- sdSubf[1,region]
				xs <- seq(from=mn.r-3*sd.r,to=mn.r+3*sd.r,length.out=1000)
				ys <- dnorm(xs,mean=mn.r,sd=sd.r)
				datNorm[[region]] <- data.frame(x=xs,y=ys)
				#p0 <- ggplot(datNorm,aes(x,y))+geom_line(color=cols[region])      
			}
			#pS[[i]] <- pS[[i]]+p0
		}
	################## End preparation #################
    subi.ls2 <- lapply(ls2,function(x) x[,i])
    in1 <- data.frame(score=subi.ls2[[1]],Input=stemName2[1])
    in2 <- data.frame(score=subi.ls2[[2]],Input=stemName2[2])
    dataFeature <- rbind(in1,in2)
    #p <- ggplot(data,aes(score,fill=Input)) + geom_histogram( aes(y = ..density.. ),alpha=1,position='dodge' ) # aes(y = ..density.. )
    #p <- p + labs(x=subNames[i])
    #if (i!=nSub) {
    #	P <- p+theme(legend.position='none')
    #} else {
    #  P <- p+theme(legend.position='top')
    #}
		if (subNames[1]!='HS') {
			if (i!=nSub) {
				pS[[i]] <- qplot(geom='blank')+geom_histogram(data=dataFeature,aes(x=score,y=..density..,fill=Input),alpha=1,position='dodge' )+geom_line(data=datNorm[['introns']],aes(x,y),color=cols['introns']) + geom_line(data=datNorm[['exons']],aes(x,y),color=cols['exons']) + geom_line(data=datNorm[['intergene']],aes(x,y),color=cols['intergene'])+labs(x=subNames[i],y='density') +theme(legend.position='none') #aes(y = ..density.. )
			} else {
				pS[[i]] <- qplot(geom='blank')+ geom_histogram(data=dataFeature,aes(x=score,y=..density..,fill=Input),alpha=1,position='dodge' ) +geom_line(data=datNorm[['introns']],aes(x,y),color=cols['introns']) + geom_line(data=datNorm[['exons']],aes(x,y),color=cols['exons']) + geom_line(data=datNorm[['intergene']],aes(x,y),color=cols['intergene']) +labs(x=subNames[i],y='density') +theme(legend.position='top')
			}
		} else {
			#if (i!=nSub) {
      #  pS[[i]] <- qplot(geom='blank') +  geom_histogram(data=dataFeature,aes(x=score,y=..density..,fill=Input),alpha=1,position='dodge' ) + labs(x=subNames[i]) +theme(legend.position='none') #aes(y = ..density.. )
      #} else {
        pS[[i]] <- qplot(geom='blank') + geom_histogram(data=dataFeature,aes(x=score,y=..density..,fill=Input),alpha=1,position='dodge' ) + labs(x=subNames[i],y='density') +theme(legend.position='top')

		}
  }
  do.call('grid.arrange',c(pS,nrow=1))
}

g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

#R_comp_ftPlot2(): plotting foreground/background distributions for one feature.
#
#INPUT infiles2: the two input bed files. e.g., c('H3K27ac.bed','H3K9me3.bed')
#INPUT subNames: the subfeatures of one feature. e.g., c('Entropy2','Entropy3','Entropy4') are good for Entropy.
R_comp_ftPlot2 <- function(ls2,infiles2,subNames=subNames0,mn=mnFeatures,sd=sdFeatures) { #itv1,itv2,   # col.ls2,c(stem1,stem2),c('Entropy2','Entropy3','Entropy4')
  library(ggplot2)
  library(gridExtra)
  nSub <- length(subNames) #ncol(ls2[[1]])
  pS <- vector('list',nSub)
  mnSubfs <- mn[subNames,,drop=F]
  sdSubfs <- sd[subNames,,drop=F]
  
	isNotHS<-subNames[1]!='HS'

	allDatFeature<-NULL
	allDatNorm<-NULL

  cols <- c('gray50','black','gray80') #introns,exons,intergene
  names(cols) <- colnames(mn)

  for (i in 1:nSub) {
    ############ Prepare xs & ys for background distributions #########
    if (isNotHS) {
      mnSubf <- mnSubfs[i,,drop=F]
      sdSubf <- sdSubfs[i,,drop=F]
      for (region in colnames(mn)) {
        mn.r <- mnSubf[1,region]
        sd.r <- sdSubf[1,region]
        xs <- seq(from=mn.r-3*sd.r,to=mn.r+3*sd.r,length.out=1000)
        ys <- dnorm(xs,mean=mn.r,sd=sd.r)
        datNorm <- data.frame(x=xs,y=ys,Category=subNames[i], Region=region)

				if(is.null(allDatNorm)){
					allDatNorm<-datNorm
				}else{
					allDatNorm<-rbind(allDatNorm, datNorm)
				}
      }
    }

    ################## End preparation #################
    subi.ls2 <- lapply(ls2,function(x) x[,i])
    in1 <- data.frame(score=subi.ls2[[1]],Input=infiles2[1])
    in2 <- data.frame(score=subi.ls2[[2]],Input=infiles2[2])
    datFeature <- rbind(in1,in2)
		datFeature$Category<-subNames[i]

		if (is.null(allDatFeature)){
			allDatFeature<-datFeature
		}else{
			allDatFeature<-rbind(allDatFeature, datFeature)
		}
  }

	g <- ggplot(allDatFeature) + geom_histogram(aes(x=score,y=..density..,fill=Input),alpha=1,position='dodge' )
	if (isNotHS) {
		g <- g + geom_line(data=allDatNorm,aes(x,y,color=Region))
	}
	g <- g + labs(x='score', y='density') + facet_wrap(.~Category, scales="free") + theme_bw() +theme(legend.position='top',strip.background=element_blank()) 
	
	print(g)
}

####################### Read in python-annotated files & derive GC% ###################
intv.ls2 <- lapply(c(options$control, options$sample),
	function(x) {
		anno <- read.delim(x,check.names=F,as.is=T)
		gc_mean <- anno[,'G%_mean']+anno[,'C%_mean']
		anno <- data.frame(anno,gc_mean=gc_mean)
		anno
	}
)

fileNames<-c(options$controlName, options$sampleName)

cmpres <- cmpr2in(intv.ls2, fileNames, options$databaseFolder, mnFeatures, sdFeatures)

```

## Catalogue of input intervals by genomic region type
The genomic region type of each input interval are identified as Intron, Exon, or Intergene (intergenic region), on account of the nature of the embedding DNA fragment. An interval can be tagged with multiple region type labels, because different region type instances may overlay and also partial overlap is counted in type identification. 
```{r compare table 1, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}
kable(cmpres$regionSum)
```

## Statistical comparison of genomic features
For all features except homology (HS), two-sample Wilcoxon test and Kolmogorov-Smirnov test are performed to compare the mean feature scores of intervals between two input sets. For homology, Zero-Inflated Poisson (ZIP) regression and Negative Binomial (NB) regression is employed to examine the difference of (median) redundancy counts of intervals between two input sets.

Med1 & Med2: median values of interval set one and interval set two.

W, W+, W-: test statistics in Wilcoxon test (for all features except HS). W, two-tailed test; W+, one-tailed, alternative hypothesis is "set two greater than set one;" W-, one-tailed, alternative "set two less than set one."

W.p, W+.p, W-.p: p-values out of Wilcoxon test.

D, D+, D-: test statistic in Kolmogorov-Smirnov test (for all features except HS). As with Wilcoxon test, "+" and "-" denote one-tailed alternative hypotheses, for "greater" and "less", respectively.

D.p, D+.p, D-.p: p-values out of Kolmogorov-Smirnov test.

*For HS, specially, W and W.p denote coefficient and p out of ZIP regression, and D and D.p denote those from NB regression.

  
```{r compare table 2, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}

testRes <- round(cmpres$cmprTbl,3)
rownames(testRes) <- gsub('gc','GC',rownames(testRes))
colnames(testRes)[c(3,4,9,10)] <- paste0(colnames(testRes)[c(3,4,9,10)],'*')
kable(testRes)
```

## Energy
Secondary structure is predicted for a 100-bp genomic window and folding energy is estimated based on thermodynamic parameters of DNA structure motifs. More information is available in SantaLucia and Hicks' review paper (doi: 10.1146/annurev.biophys.32.110601.141800).  Energy value is first calculated for each base site on account of the enclosing 100-bp window, and then averaged across each interval.

```{r compare Energy, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}
############ Energy ###################
subNames <- c('delta_G37','delta_H')
col.ls2 <- lapply(intv.ls2,function(x,cols) x[,cols,drop=F], paste(subNames,'mean',sep='_') )
R_comp_ftPlot2(col.ls2,fileNames,subNames)
```

## Entropy
For a base site, Shannon entropy is employed to quantify the variability of n-mer oligonucleotides within a 100-bp enclosing window centered upon the site in question, and the base-wise entropy values are averaged to an interval-wise entropy. 2-mer, 3-mer, and 4-mer oligonucleotides are separately considered, thus giving rise to Entropy2, Entropy3, and Entropy4. 

```{r compare Entropy, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}
############ Entropy ##################
subNames <- c(paste0('Entropy',2:4))
col.ls2 <- lapply(intv.ls2,function(x,cols) x[,cols,drop=F], paste(subNames,'mean',sep='_') )
R_comp_ftPlot2(col.ls2,fileNames,subNames)
```

## GC content
For a base site, GC content is defined as the number of guanosines (G) and cytidines (C) in the 100-bp enclosing window centered upon the site in question. The GC content of an interval is the mean GC content across all bases in the interval. 

```{r compare GC, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}
########## GC #################
subNames <- c('gc')
col.ls2 <- lapply(intv.ls2,function(x,cols) x[,cols,drop=F], paste(subNames,'mean',sep='_') )
R_comp_ftPlot2(col.ls2,fileNames,subNames)
```

## Homology score (redundancy count)
For a base site, homology score is defined as the incidence of perfect or near-perfect (1 base mismatch) alignment to the whole genome of the 100-bp enclosing window centered upon the site in question. The homology score of an interval is the median mapping count across all bases in the interval. 

```{r compare HS, echo=F, comment=F, warning=FALSE, dependson=F, message=FALSE}
########## HS #################
subNames <- c('HS')
col.ls2 <- lapply(intv.ls2,function(x,cols) x[,cols,drop=F], paste(subNames,'mean',sep='_') )
R_comp_ftPlot2(col.ls2,fileNames,subNames)
```

